The Radio Sky at Meter Wavelengths: $m$-Mode Analysis Imaging with the Owens Valley Long Wavelength Array

### 2017-06-26 paper.tex
s/OVRO LWA/OVRO-LWA

% OUTLINE
% =======
%
% 1. Introduction
%
%   Here I will introduce the need for a new high-fidelity, low-frequency map of the sky. The
%   justifications are: 21 cm foreground mapping, galactic structure (?), others?
%
% 2. Observations
%
%   2.1 - Introduce the Owens Valley LWA
%   2.2 - Describe the 100 hour dataset
%   2.3 - Calibration strategy
%   2.4 - Source removal strategy
%   2.5 - Horizon RFI suppression strategy
%   2.6 - Beam fitting work
%   2.7 - Ionospheric conditions
%
% 3. Imaging
%
%   Give a brief overview of other widefield imaging algorithms. This should set the context for the
%   advantages of m-mode analysis.
%
%   3.1 - Introduce m-mode analysis imaging
%   3.2 - Describe my implementation of m-mode analysis on the ASTM
%   3.3 - L-curves (?)
%   3.4 - CLEAN
%
% 4. Results
%
%   Present sky maps with an estimate of the thermal noise in each pixel. Systematic errors will be
%   described in the following sub-sections.
%
%   4.1 - Horizon RFI
%   4.2 - Ionospheric effects
%   4.3 - Beam errors
%   4.4 - Peeling failures
%
% 5. Discussion
%
%   Compare with the Haslam map and Jayce's map.
%
% 6. Conclusion
%
% FIGURES
% =======
%
% * Antenna layout of the OVRO LWA
% * Before/after snapshot image of the RFI removal
% * Image of the empirical beam (with source tracks)
% * Plot of ionospheric scintillation and refraction of an A-team source
% * Example L-curve
% * Zoom-in of the dirty PSF
% * Sky maps with their boot-strapped thermal noise estimates
% * (Model) image of the horizontal stripes that show up if the horizon RFI is not removed
% * (Model) image of the PSF distortion from ionospheric effects (scintillation, refraction)
% * (Model) image of the systematic errors in the map created from beam errors
% * (Model) image of the errors created from sporadic peeling failures

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{figures/antenna-layout}
    \caption{
        This figure shows the antenna layout for the OVRO LWA. Black dots correspond to antennas
        within the 200 m diameter core of the array. The 32 open circles are the expansion antennas
        built in early 2016 in order to increase the longest baseline to 1.5 km. The 5 crosses are
        antennas equipped with noise-switched front ends.
    }
    \label{fig:antenna-layout}
\end{figure}

The Owens Valley Long Wavelength Array (OVRO LWA) is a 288-element interferometer located at the
Owens Valley Radio Observatory (OVRO) near Big Pine, California.  The OVRO LWA is a low-frequency
instrument with instantaneous bandwidth covering 27.384 MHz to 84.912 MHz and 24 kHz channelization.
Each antenna stand hosts two perpendicular broadband dipoles so that there are $288\times2$ signal
paths in total. These signal paths feed into the 512-input LEDA correlator
\citep{2015JAI.....450003K}, which allows the OVRO LWA to capture the entire visible hemisphere in a
single snapshot image.  In the current configuration 32 antennas (64 signal paths) are unused.

@@

Beginning at 2017-02-17 12:00:00 UTC time, 28 consecutive hours of data was collected. This time was
chosen based on the fact that it was raining at OVRO and rain tends to improve the low-frequency RFI
environment considerably. During this time the OVRO LWA operated as a zenith-pointing drift scanning
interferometer.  The correlator dump time was selected to be 13 seconds such that the correlator
output evenly divides a sidereal day.

@@

The antenna complex gains are updated once per day from a 9 hour track of data when Cas A and Cyg A
are at high elevations.  These two sources are -- by an order of magnitude -- the brightest radio
sources in the northern hemisphere.  Therefore the optimal time to solve for the interferometer's
gain calibration is when Cas A and Cyg A are at high elevations. A calibration obtained in this
manner is good consistent with the following day's calibration to within a 1\% error in the flux
scale and 1$'$ error in the astrometry \todo{double check these numbers}. These residual errors are
likely driven by differing ionospheric conditions.

@@

In order to generate wide-field images of the sky, the response of the antenna to the sky must be
known. Fortunately drift-scanning interferometers like the OVRO LWA can empirically measure their
primary beam under a mild set of symmetry assumptions \citep{2012AJ....143...53P}. In this work we
assume that the primary beam is invariant under north-south and east-west flips, and additionally
that the $x$- and $y$-dipoles have the same response to the sky after rotating one by 90$^\circ$.
We measure the flux of several point sources \todo{which ones??} as they pass through the sky and
then fit a beam model composed of Zernike polynomials to those flux measurements. We select the
basis functions to have the desired symmetry.

@@

The largest angular scale of Cas A is $\sim$8 arcminutes, while the largest angular scale of Cyg A
is $\sim$2 arcminutes. With a 10 arcminute resolution, the OVRO LWA marginally resolves both
sources. A resolved source model is needed for both sources. We fit a self-consistent resolved
source model to eachs ource. This is performed by minimizing the minimizing the variance within an
aperture located on each source after peeling. By phasing up a large number of integrations before
imaging (at least over 1 hour) it is possible to smear out the contribution of the rest of the sky.
We then use NLopt's Sbplx routine \citep{nlopt, sbplx} to vary the parameters in a source model
until the variance within the aperture is minimized.

@@

The Sun can be trivially removed from any map of the sky by constructing the map using only data
taken at night. A map of the entire sky can be obtained by using observations spaced 6 months apart.
However the OVRO LWA 100 hour dataset consists of 100 consecutive hours of data, and hence roughly
half of the data is from the day.

@@

The Owens Valley is an important source of water and power for the city of Los Angeles.
Unfortunately this means that high voltage power lines run along the valley to the west of the OVRO
LWA. Some of these power line poles have faulty insulators that arc and produce pulsed, broadband
RFI. Because these poles exist in the near-field of the array, we have been able to localize some of
them by using the curvature of the incoming wavefront. Efforts are currently underway to work with
the utility pole owners to have these insulators replaced.

@@

In this context we will define a transit telescope as any interferometer where the response pattern
of the individual elements does not change with respect to time. This may be an interferometer like
the OVRO LWA where the correlation elements are fixed dipoles, but it may also be an interferometer
like the MWA if the steerable beams are held in a fixed position (not necessarily at zenith).

@@

In practice we now need to pick the set of spherical harmonics we will use to represent the sky. For
an interferometer like the OVRO LWA with lots of short baselines, a sensible choice is to use all
spherical harmonics with $l\le l_\text{max}$ for some $l_\text{max}$. The parameter $l_\text{max}$
is determined by the maximum baseline length of the interferometer.  For an interferometer without
short spacings, a minimum value for $l$ might also be used. This $l_\text{min}$ parameter should be
determined by the minimum baseline length.  When creating the maps presented in this paper, we use
$l_\text{min} = 0$ and $l_\text{max} = 1000$.

@@

The size of a typical block in the transfer matrix is
$(2N_\text{base}N_\text{freq})\times(l_\text{max})$. If each element of the matrix is stored as a
64-bit complex floating point number, a single block is 500 MB for the case of single-channel
imaging with the OVRO LWA. Compare this number with the 5 PB required for the naive approach.  The
power of $m$-mode analysis is the block-diagonal structure of
equation~\ref{eq:m-mode-matrix-equation}.  By breaking up the problem into $N$ independent blocks,
the computational complexity involved in inverting the equation is reduced by a factor $N^3$. For
the case of the OVRO LWA the equation breaks up into $\sim10^3$ blocks and so we save a factor of
$\sim10^9$ in processing time by using $m$-mode analysis.

@@

\subsection{$m$-Mode Analysis with the OVRO LWA}

>>>>

% OUTLINE
% =======
%
% 1. Introduction
%
%   Here I will introduce the need for a new high-fidelity, low-frequency map of the sky. The
%   justifications are: 21 cm foreground mapping, galactic structure (?), others?
%
% 2. Observations
%
%   2.1 - Introduce the Owens Valley LWA
%   2.2 - Describe the 100 hour dataset
%   2.3 - Calibration strategy
%   2.4 - Source removal strategy
%   2.5 - Horizon RFI suppression strategy
%   2.6 - Beam fitting work
%   2.7 - Ionospheric conditions
%
% 3. Imaging
%
%   Give a brief overview of other widefield imaging algorithms. This should set the context for the
%   advantages of m-mode analysis.
%
%   3.1 - Introduce m-mode analysis imaging
%   3.2 - Describe my implementation of m-mode analysis on the ASTM
%   3.3 - L-curves (?)
%   3.4 - CLEAN
%
% 4. Results
%
%   Present sky maps with an estimate of the thermal noise in each pixel. Systematic errors will be
%   described in the following sub-sections.
%
%   4.1 - Horizon RFI
%   4.2 - Ionospheric effects
%   4.3 - Beam errors
%   4.4 - Peeling failures
%
% 5. Discussion
%
%   Compare with the Haslam map and Jayce's map.
%
% 6. Conclusion
%
% FIGURES
% =======
%
% * Antenna layout of the OVRO-LWA
% * Before/after snapshot image of the RFI removal
% * Image of the empirical beam (with source tracks)
% * Plot of ionospheric scintillation and refraction of an A-team source
% * Example L-curve
% * Zoom-in of the dirty PSF
% * Sky maps with their boot-strapped thermal noise estimates
% * (Model) image of the horizontal stripes that show up if the horizon RFI is not removed
% * (Model) image of the PSF distortion from ionospheric effects (scintillation, refraction)
% * (Model) image of the systematic errors in the map created from beam errors
% * (Model) image of the errors created from sporadic peeling failures

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{figures/antenna-layout}
    \caption{
        This figure shows the antenna layout for the OVRO-LWA. Black dots correspond to antennas
        within the 200 m diameter core of the array. The 32 open circles are the expansion antennas
        built in early 2016 in order to increase the longest baseline to 1.5 km. The 5 crosses are
        antennas equipped with noise-switched front ends.
    }
    \label{fig:antenna-layout}
\end{figure}

The Owens Valley Long Wavelength Array (OVRO-LWA) is a 288-element interferometer located at the
Owens Valley Radio Observatory (OVRO) near Big Pine, California.  The OVRO-LWA is a low-frequency
instrument with instantaneous bandwidth covering 27.384 MHz to 84.912 MHz and 24 kHz channelization.
Each antenna stand hosts two perpendicular broadband dipoles so that there are $288\times2$ signal
paths in total. These signal paths feed into the 512-input LEDA correlator
\citep{2015JAI.....450003K}, which allows the OVRO-LWA to capture the entire visible hemisphere in a
single snapshot image.  In the current configuration 32 antennas (64 signal paths) are unused.

@@

Beginning at 2017-02-17 12:00:00 UTC time, 28 consecutive hours of data was collected. This time was
chosen based on the fact that it was raining at OVRO and rain tends to improve the low-frequency RFI
environment considerably. During this time the OVRO-LWA operated as a zenith-pointing drift scanning
interferometer.  The correlator dump time was selected to be 13 seconds such that the correlator
output evenly divides a sidereal day.

@@

The antenna complex gains are updated once per day from a 9 hour track of data when Cas A and Cyg A
are at high elevations \todo{the rainy day dataset actually uses a shorter track of data, we should
try using the 9 hour track again}.  These two sources are -- by an order of magnitude -- the
brightest radio sources in the northern hemisphere.  Therefore the optimal time to solve for the
interferometer's gain calibration is when Cas A and Cyg A are at high elevations. A calibration
obtained in this manner is good consistent with the following day's calibration to within a 1\%
error in the flux scale and 1$'$ error in the astrometry \todo{double check these numbers}. These
residual errors are likely driven by differing ionospheric conditions.

@@

In order to generate wide-field images of the sky, the response of the antenna to the sky must be
known. Fortunately drift-scanning interferometers like the OVRO-LWA can empirically measure their
primary beam under a mild set of symmetry assumptions \citep{2012AJ....143...53P}. In this work we
assume that the primary beam is invariant under north-south and east-west flips, and additionally
that the $x$- and $y$-dipoles have the same response to the sky after rotating one by 90$^\circ$.
We measure the flux of several point sources \todo{which ones??} as they pass through the sky and
then fit a beam model composed of Zernike polynomials to those flux measurements. We select the
basis functions to have the desired symmetry.

@@

The largest angular scale of Cas A is $\sim$8 arcminutes, while the largest angular scale of Cyg A
is $\sim$2 arcminutes. With a 10 arcminute resolution, the OVRO-LWA marginally resolves both
sources. A resolved source model is needed for both sources. We fit a self-consistent resolved
source model to eachs ource. This is performed by minimizing the minimizing the variance within an
aperture located on each source after peeling. By phasing up a large number of integrations before
imaging (at least over 1 hour) it is possible to smear out the contribution of the rest of the sky.
We then use NLopt's Sbplx routine \citep{nlopt, sbplx} to vary the parameters in a source model
until the variance within the aperture is minimized.

@@

The Sun can be trivially removed from any map of the sky by constructing the map using only data
taken at night. A map of the entire sky can be obtained by using observations spaced 6 months apart.
However the OVRO-LWA 100 hour dataset consists of 100 consecutive hours of data, and hence roughly
half of the data is from the day.

@@

The Owens Valley is an important source of water and power for the city of Los Angeles.
Unfortunately this means that high voltage power lines run along the valley to the west of the OVRO-
LWA. Some of these power line poles have faulty insulators that arc and produce pulsed, broadband
RFI. Because these poles exist in the near-field of the array, we have been able to localize some of
them by using the curvature of the incoming wavefront. Efforts are currently underway to work with
the utility pole owners to have these insulators replaced.

@@

In this context we will define a transit telescope as any interferometer where the response pattern
of the individual elements does not change with respect to time. This may be an interferometer like
the OVRO-LWA where the correlation elements are fixed dipoles, but it may also be an interferometer
like the MWA if the steerable beams are held in a fixed position (not necessarily at zenith).

@@

In practice we now need to pick the set of spherical harmonics we will use to represent the sky. For
an interferometer like the OVRO-LWA with lots of short baselines, a sensible choice is to use all
spherical harmonics with $l\le l_\text{max}$ for some $l_\text{max}$. The parameter $l_\text{max}$
is determined by the maximum baseline length of the interferometer.  For an interferometer without
short spacings, a minimum value for $l$ might also be used. This $l_\text{min}$ parameter should be
determined by the minimum baseline length.  When creating the maps presented in this paper, we use
$l_\text{min} = 0$ and $l_\text{max} = 1000$.

@@

The size of a typical block in the transfer matrix is
$(2N_\text{base}N_\text{freq})\times(l_\text{max})$. If each element of the matrix is stored as a
64-bit complex floating point number, a single block is 500 MB for the case of single-channel
imaging with the OVRO-LWA. Compare this number with the 5 PB required for the naive approach.  The
power of $m$-mode analysis is the block-diagonal structure of
equation~\ref{eq:m-mode-matrix-equation}.  By breaking up the problem into $N$ independent blocks,
the computational complexity involved in inverting the equation is reduced by a factor $N^3$. For
the case of the OVRO-LWA the equation breaks up into $\sim10^3$ blocks and so we save a factor of
$\sim10^9$ in processing time by using $m$-mode analysis.

@@

\subsection{$m$-Mode Analysis with the OVRO-LWA}
